---
title: "STAT 7510 - Exam 2 Take Home - Question 2"
author: "Benjamin Tjepkes"
date: "2024-07-23"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    toc_depth: 3
    number_sections: no
    number_offset: 1
    theme: lumen
    font_size: 16
---

<style type="text/css">
  body{
  font-size: 14pt;
}

  h2 {
  font-size: 24pt;
  color: #e7700d;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# Load in CITYCRIME data
Y <- read.table(file = "./T15_1_CITYCRIME.dat")

# Rename columns
names(Y) <- c("City", "Murder", "Rape", "Robbery", "Assault", "Burglary", "Larceny", "AutoTheft")

# Specify row names
row.names(Y) <- Y[ , "City"]

# View structure of data
dplyr::glimpse(Y)
```

## A

**Fit a Kmeans model to the data with k=4 and nstart=20**

```{r}
# All variables are crime rates per 100,000 population, so not scaled
# Subset Y to exclude city name
Y_clean <- Y[, 2:8]

# Set seed for reproducibility
set.seed(1)

# Fit a k-means model
km.fit <- kmeans(x = Y_clean, centers = 4, nstart = 20)

# Print total within cluster sum of squares
km.fit$tot.withinss
```


## B

**Fit 3 Hierarchical Clustering models with method = ”complete”, method = ”average”, and method = ”single”.**

```{r}
# Perform H-Cluster with "complete"
hc.complete <- hclust(d = dist(Y_clean), method = "complete")
  
# Perform H-Cluster with "average"
hc.average <- hclust(d = dist(Y_clean), method = "average")

# Perform H-Cluster with "single"
hc.single <- hclust(d = dist(Y_clean), method = "single")
```


## C

**Plot the 3 Hierarchical Clustering models and compare.**

```{r}

# Plot HC complete dendrogram
plot(hc.complete, main="Complete Linkage", xlab = "", sub = "", cex = 0.9)

# Plot HC average dendrogram
plot(hc.average, main = "Average Linkage", xlab = "", sub = "", cex = 0.9)

# Plot HC single dendrogram
plot(hc.single,main="Single Linkage", xlab = "", sub = "", cex = 0.9)
```


The "complete" and "average" linkages result in practically the same clusters and branching. The "single" linkages results in a very bushy dendrogram with more dissimilarity between groups.


## D

**Which of the 3 Hierarchical Clustering models is closest to the k-means model from part (a). You will need the cutree function.**

```{r}
# Cut complete linkage dendrogram
hc.complete.cut <- cutree(tree = hc.complete, k = 4)

# Compare complete hierarchical to k-means
table(km.fit$cluster, hc.complete.cut)
```

```{r}
# Cut average linkage dendrogram
hc.average.cut <- cutree(tree = hc.average, k = 4)

# Compare average hierarchical to k-means
table(km.fit$cluster, hc.average.cut)
```

```{r}
# Cut single linkage dendrogram
hc.single.cut <- cutree(tree = hc.single, k = 4)

# Compare single hierarchical to k-means
table(km.fit$cluster, hc.single.cut)
```

Based on the above contingency tables, it looks like the "complete" linkage hierarchical clustering method has the highest agreement with the k-means clustering method.


## Session Info

```{r}
sessionInfo()
```

